\documentclass[xcolor={table,usenames,dvipsnames}]{beamer}

% Suggested improvements:
%  - Actually apply AllMinAs to the example in a step-by-step manner.
%  - Remove references, they are too noisy.
%  - Expand on Complexity.
%  - How is this useful to the person working with the ontology? (maybe mention MaxAs)
%  - Use overlays consistently throughout the presentation.
%  - Maybe explain the algorithms using text and bullet lists with overlays.

\usetheme[sectionpage=progressbar,subsectionpage=progressbar,block=fill]{metropolis}

\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{mathrsfs}
\usepackage{bibentry}
\usepackage[scale=1.5]{ccicons}

\usepackage{hyperref}

\usepackage{amsmath,amssymb} % For \flalign, \top, ...
\usepackage[algoruled,linesnumbered]{algorithm2e}

\setlength{\algoheightrule}{0.5pt} % thickness of the rules above and below
\setlength{\algotitleheightrule}{0.5pt} % thicknes of the rule below the title

\usepackage{nicefrac,xspace,csquotes}

% If we want article-style math even in the presentation.
%\usefonttheme[onlymath]{serif}

\newcommand{\elp}{\ensuremath{\mathcal{EL^+}}\xspace}
\newcommand{\elpp}{\ensuremath{\mathcal{EL^{++}}}\xspace}
\newcommand{\el}{\ensuremath{\mathcal{EL}}\xspace}
\newcommand{\hl}{\ensuremath{\mathcal{HL}}\xspace}
\newcommand{\tb}{\ensuremath{\mathcal{T}}\xspace} % TBox
\newcommand{\nc}{\ensuremath{\mathsf{N_C}}\xspace} % concept names
\newcommand{\nr}{\ensuremath{\mathsf{N_R}}\xspace} % role names

\newcommand{\snomed}{\textsc{Snomed CT}\xspace}
\newcommand{\galen}{\textsc{Galen}\xspace}

\newcommand{\subsume}{\sqsubseteq}

\title{Pinpointing in the Description Logic \texorpdfstring{\elp}{EL+}}
\author{\texorpdfstring{Lorenz Leutgeb \newline \texttt{\footnotesize lorenz@leutgeb.xyz}}{Lorenz Leutgeb <lorenz@leutgeb.xyz>}}
\institute{International Center for Computational Logic, TU Dresden\\[10mm] \begin{center}\ccbysa \\[1mm] \tiny{This work is licensed under \href{http://creativecommons.org/licenses/by-sa/4.0/}{Creative Commons BY-SA 4.0}.}\end{center}}
\date{2017-01-18}

\nobibliography*

\begin{document}
 
\begin{frame}[plain]
\maketitle
\end{frame}

\begin{frame}[plain]
Paper available for download at:\\[3mm] \url{https://lorenz.leutgeb.xyz/paper/elp.pdf}\vfill

\noindent\rule{\textwidth}{0.5pt}\vfill

Based on:\\[3mm]

\large{Pinpointing in the Description Logic \elp}\\[1mm]
Baader, Pe\铨犰嵇岈犷吁铘轶蜷鲠蜥痫蝾苘鄄眄嘲荇屮趔躔弪筱蜷痿翳令铛犰清蝽犷蔑铈弪孱沐镱辽舶胺苘鄢眄荏磲祆茔镬矧珧狴ㄤ弭衢戾蝈驽蝈钽轭翳孱洎苠钿骝犴妪茆彗轭骝犴妪羚孱溽荇徕戾镦泔铘孱趔苠钿骝犴妪荏邈糸镱深趄镤蹉糸镱荏踱箦泗轱铥御铘狲犷渝磲铘殂簖茆彗轭骝犴妪御铘狲犷渝磲铘殂茔轸遨渝惝铂遭飚陛矧殓昌荏磲祆茯秣泔祜蝮昌硖殓梏买秣睢舶麒轸妪荏弭戾铉翳荇徕泔祗屦仇酏茆彗轭翎怩灬螨禳沣沣泓荑扉铄吾礤御铘狲渝磲铘殂荑苠苠祓苘荑扉铄荑扉铄燥ぼ麸黏ぼ腻祠徂嗓ぼ怩祆弭ぼ怩祆弭ぼ怩祆弭苘蔑铌躅泗轱荏胥狃っ奚茔狃霓嗓ぼ怩祆弭ぼ怩祆弭ぼ怩祆弭苘砒轶翦铘獒义篝虍苠轶趔ぼ狍簸ぼ怩祆弭ぼ怩祆弭苘荑扉铄荑扉铄敲绍骘雉铒翦清铄蜥蔑钽屦深沆躞轱铨っ荏耋踱箦翦膜っ奚荏踱箦翦霓嗓ぼ怩祆弭ぼ怩祆弭ぼ怩祆弭苘蔑钽屦腻骈铋糸镱っ苠聃轹膜っ奚霓嗓ぼ怩祆弭ぼ怩祆弭ぼ怩祆弭苘绎戾深沆躞轱を弑茔轵茔滹趔茔轵蜻荏耋踱箦翦螭を弑奚茔轵茔滹趔茔轵蜻钷荏踱箦翦筠嗓ぼ怩祆弭苘荑扉铄苠钿翎怩灬螨ぼ翦酐ぼ狍簸茺荛苣屐翎奚茼殇荇屮酐翳弪屮轶趔荛苣屐翎奚荇屮酐螽舢 (x, y) \in r^I \text{ and } y \in C^I \}$
\begin{itemize}
\item{Concept Descriptions $C, D$ (inductively)}
\item{Role Names $r_1, \ldots, r_n, s$}
\item{Classical Interpretation $I = (\Delta^I, \cdot^I)$}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Example: \hl and Horn Logic Programming (cf.\cite[Sec.2]{orig2})}

\begin{align*}
	\mathsf{woman} &\subsume \mathsf{person} & \text{\texttt{person :- }}&\text{\texttt{woman.}}\\
	\mathsf{man} &\subsume \mathsf{person} & \text{\texttt{person :- }}&\text{\texttt{man.}}\\
	\mathsf{parent} \sqcap \mathsf{woman} &\subsume \mathsf{mother} & \text{\texttt{mother :- }}&\text{\texttt{parent, woman.}}
\end{align*}

\end{frame}

\subsection{TBoxes and Concept Subsumption}

\begin{frame}{TBoxes and Concept Subsumption}
\begin{itemize}
\item We consider knowledge bases that are  \emph{finite sets of axioms}, called \alert{TBoxes}, denoted $\tb$.
\item Key questions wrt.TBoxes are \alert{satisfiability} and \alert{concept subsumption}.
\end{itemize}

\begin{definition}[{Concept Subsumption, cf.\cite[Def.1]{orig1}}]
\label{def:subs}
Given two concept descriptions $C, D$ and a TBox \tb, $C$ is subsumed by $D$ wrt.\tb (written $C \sqsubseteq_\tb D$) if for every interpretation $I$ that satisfies \tb we have $C^I \subseteq D^I$.
\end{definition}
\end{frame}

\subsection{Pinpointing}

\begin{frame}{Motivation for Pinpointing}
\alert<1>{Given a TBox \ldots}
{\footnotesize \begin{flalign}
\mathsf{Human} &\subsume \exists \mathsf{parent} . \mathsf{Human} & \text{Humans have a human parent.} \tag{$a_1$} \\
\mathsf{Human} &\subsume \mathsf{Monkey} & \text{Humans are monkeys.} \tag{$a_2$} \\
\exists \mathsf{parent} . \mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkey parent? It's an animal.} \tag{$a_3$} \\
\mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkeys are animals.} \tag{$a_4$} \\
\mathsf{Fish} &\subsume \mathsf{Animal} & \text{Fish are animals.} \tag{$a_5$}
\end{flalign}}

\visible<2->{
\alert<2>{We observe that \ldots}
\begin{flalign*}
\mathsf{Human} \subsume \mathsf{Animal}
\end{flalign*}}

\visible<3->{\alert<3>{And ask ourselves: Why?}}
\end{frame}

\begin{frame}{Motivation for Pinpointing}
Formalization of our question yields \alert{Pinpointing}, a process that results in \emph{minimal axiom sets}:\\[5mm]

\begin{definition}[{MinA, cf.\cite[Def.2 without partitioning $\tb$]{orig1}}]
\label{def:mina}
Let $\tb$ be a TBox and $A, B$ concept names occurring in it such that $A \subsume_\tb B$. Then a \alert{minimal axiom set (MinA)} for \tb wr.t.$A \subsume B$ is a subset $\mathcal{S} \subseteq \tb$ such that $$A \subsume_\mathcal{S} B$$ and for all $\mathcal{S'} \subset \mathcal{S}$ we have $$A \not \subsume_\mathcal{S'} B$$
\end{definition}
\end{frame}

\begin{frame}{Motivation for Pinpointing}

{\footnotesize
\begin{flalign}
\mathsf{Human} &\subsume \exists \mathsf{parent} . \mathsf{Human} & \text{Humans have a human parent.} \tag{$a_1$} \\
\mathsf{Human} &\subsume \mathsf{Monkey} & \text{Humans are monkeys.} \tag{$a_2$} \\
\exists \mathsf{parent} . \mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkey parent? It's an animal.} \tag{$a_3$} \\
\mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkeys are animals.} \tag{$a_4$} \\
\mathsf{Fish} &\subsume \mathsf{Animal} & \text{Fish are animals.} \tag{$a_5$}
\end{flalign}}

$$\mathsf{Human} \subsume \mathsf{Animal}$$

$$ \text{Minimal Axiom Sets (MinAs):  } \{ a_2, a_4 \}, \{  a_1, a_2, a_3 \}$$
\end{frame}

\begin{frame}{Example: \snomed \cite[Fig.6.8, p.128]{meng-phd}}

{\footnotesize{
\begin{flalign*}
\textsf{direct-procedure-site} \subsume & \textsf{procedure-site} \\
AmputationOfFinger \subsume & AmputationOfFingerNotThumb \\
\textsf{AmputationOfFingerNotThumb} \equiv & \textsf{HandExcision} \sqcap \\ & \exists roleGroup . ( \\ & \quad  \exists\textsf{direct-procedure-site} . \textsf{Finger}_S \sqcap  \\ & \quad \exists \textsf{method} . \textsf{Amputation} ) \\
\textsf{AmputationOfHand} \equiv & \textsf{HandExcision} \sqcap \\ & \exists roleGroup . ( \\ & \quad  \exists\textsf{direct-procedure-site} . \textsf{Finger}_S \sqcap  \\ & \quad \exists \textsf{method} . \textsf{Amputation} ) \\
Finger_S \subsume & DigitOfHand_S \sqcap Hand_P \\
Hand_P \subsume & Hand_S \sqcap UpperExtremity_P
\end{flalign*}
}}
\end{frame}

\section{Algorithms}

\begin{frame}{White-Box vs.Black-Box Algorithms}
\begin{description}
\item[White-Box] Inspects syntax of axioms, more \enquote{low-level}.\\[4mm]
\item[Black-Box] Relies on reasoning services (subsumption) only.
\end{description}
\end{frame}

\subsection{Pinpointing via Labeling}

\begin{frame}{Subsumption Algorithm for \elp}
Based on following \alert{completion rules}\footnote{adapted from \cite[Fig.1]{orig1,orig2}, see also \cite[Fig.5.2, p.104]{meng-phd}} wrt.a TBox $\tb$. Rule $i$ is applicable if $a_i \in \tb$ and $P_i \subseteq \tb' \setminus \tb$. If rule $i$ is applied, then $q_i$ is added to $\tb'$.
\begin{table}
\begin{center}
\rowcolors{2}{mLightBrown!20}{white}
\setlength\tabcolsep{4pt}
\begin{tabular}{|c|rcll|l|}
\hline
{\bfseries } & \multicolumn{4}{c|}{\bfseries Applicability} & \multicolumn{1}{c|}{\bfseries Result} \\
$i$
 & \multicolumn{3}{c}{$a_i$ (axiom)} & $P_i$ (set of $\tb$-seq) & $q_i$ ($\tb$-seq) \\
\hline\hline
1& $A_1 \sqcap \cdots \sqcap A_n$&$\sqsubseteq$&$B$ & $X \subsume A_1 , \ldots , X \subsume A_n$ & $X \subsume B$ \\
2& $A$&$\subsume$&$\exists r.B$ & $X \subsume A$ & $X \subsume \exists r . B$ \\
3& $\exists r.A$&$\subsume$&$B$ & $X \subsume \exists r . Y, Y \subsume A$ & $X \subsume B$ \\
4& $r$&$\subsume$&$s$ & $X \subsume \exists r . Y$ & $X \subsume \exists s . Y$ \\
5& $r \circ r'$&$\subsume$&$s$ & $ X \subsume \exists r . Y, Y \subsume \exists r'. Z$ & $X \subsume \exists s . Z$\\
\hline
\end{tabular}
\end{center}
\label{tbl:rules}
\end{table}
\end{frame}

\begin{frame}
\begin{algorithm}[H]
  \KwIn{An \elp TBox $\tb$ in normal form over $\nc$ and $A, B \in \nc$.}
  \KwOut{\enquote{yes} if $A \subsume_\tb B$ holds, \enquote{no} otherwise.}
  $\tb' := \{ A \subsume A, A \subsume \top \mid A \in \nc \}$ \\
  \While{there is a rule $i$ s.t. $1 \leq i \leq 5$, $a_i \in \tb$ and $P_i \subseteq \tb'$}{
    $\tb' := \tb' \cup \{ q_i \}$ \\
  }
  \Return{\emph{\enquote{yes}} \emph{\textbf{if}} $A \subsume B \in \tb'$ \emph{\textbf{otherwise}} \emph{\enquote{no}}}
  \caption{\textsc{Subsumption}($\tb$, $A$, $B$)}
  \label{alg:sub}
\end{algorithm}
\end{frame}

\begin{frame}{Subsumption Algorithm for \elp}
\alert{The algorithm \ldots}
\begin{itemize}[<+->]
\item requires a normalized TBox. Normalization is always possible and can be computed in linear time\footnote{proof for $\el$ in \cite[Lemma 6.2]{book} and \elpp in \cite[Lemma 1]{pushing}}.
\item restricts to concept names. We may introduce new concept names $A \subsume C, D \subsume B$ for any concepts $C, D$.
\item is correct\footnotemark \footnotetext{\label{mynote} \cite[Thm.1]{orig2}}: $A \subsume_\tb B \text{ iff }A \subsume B \in \tb'$
\item runs in time polynomial in the size of the input TBox\footnotemark[4].
\item actually computes all concept subsumptions. This is easily extended to a \emph{classification algorithm}.
\end{itemize}
\end{frame}

\begin{frame}{Pinpointing Formula}

{\footnotesize
\begin{flalign}
\mathsf{Human} &\subsume \exists \mathsf{parent} . \mathsf{Human} & \text{Humans have a human parent.} \tag{$a_1$} \\
\mathsf{Human} &\subsume \mathsf{Monkey} & \text{Humans are monkeys.} \tag{$a_2$} \\
\exists \mathsf{parent} . \mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkey parent? It's an animal.} \tag{$a_3$} \\
\mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkeys are animals.} \tag{$a_4$} \\
\mathsf{Fish} &\subsume \mathsf{Animal} & \text{Fish are animals.} \tag{$a_5$}
\end{flalign}
}
$$\mathsf{Human} \subsume \mathsf{Animal}$$

$$ \text{Minimal Axiom Sets (MinAs):  } \{ a_2, a_4 \}, \{  a_1, a_2, a_3 \}$$
\end{frame}

\begin{frame}{Pinpointing Formula}

{\footnotesize
\begin{flalign}
\mathsf{Human} &\subsume \exists \mathsf{parent} . \mathsf{Human} & \text{Humans have a human parent.} \tag{$a_1$} \\
\mathsf{Human} &\subsume \mathsf{Monkey} & \text{Humans are monkeys.} \tag{$a_2$} \\
\exists \mathsf{parent} . \mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkey parent? It's an animal.} \tag{$a_3$} \\
\mathsf{Monkey} &\subsume \mathsf{Animal} & \text{Monkeys are animals.} \tag{$a_4$} \\
\mathsf{Fish} &\subsume \mathsf{Animal} & \text{Fish are animals.} \tag{$a_5$}
\end{flalign}
}
$$\mathsf{Human} \subsume \mathsf{Animal}$$

$$ \text{Pinpointing Formula:  } a_2 \wedge (a_4 \vee (a_1 \wedge a_3))$$
\end{frame}

\begin{frame}{Pinpointing Formula}

\begin{itemize}
\item Let $lab(\tb)$ be the set of labels of all axioms in \tb.
\item Let $\mathcal{V} \subseteq lab(\tb)$ be a valuation wrt.\tb.
\item Let $\tb_\mathcal{V} = \{ a \in \tb \mid lab(a) \in \mathcal{V}\}$ be the selection of axioms with a label that is true under $\mathcal{V}$.\\[6mm]
\end{itemize}

\begin{definition}[{Pinpointing Formula \cite[Def.3]{orig1}}]
Given an \elp TBox \tb and concept names $A,B$ occurring in it, a monotone Boolean formula $\psi$ over $lab(\tb)$ is a \alert{pinpointing formula} for $\tb$ wrt.$A \subsume B$ if for every valuation $\mathcal{V} \subseteq lab(\tb)$ it holds that $A \subsume_{\tb_\mathcal{V}} B$ iff $\mathcal{V}$ satisfies $\psi$.
\end{definition}

\begin{itemize}
\item Given a pinpointing formula $\psi$, we can construct corresponding MinAs \cite[Prop.1]{orig2}:
$$\{ \tb_{\mathcal{V}} \mid \mathcal{V} \models \psi \text{ and } \mathcal{V} \text{ is } \text{$\subseteq$-minimal} \}$$
\end{itemize}
\end{frame}

\begin{frame}
\vspace{3mm}
\small{
\begin{algorithm}[H]
  \KwIn{An \elp TBox $\tb$ in normal form over $\nc$.}
  \KwOut{A TBox $\tb'$ and a labeling function.}
  Assign $\tb' := \{A \subsume A, A \subsume \top \mid A \in \nc \}$ and $lab(a) := \texttt{true}$ f.a.$a \in \tb'$\\
  \While{there is a rule $i$ s.t. $1 \leq i \leq 5$, $a_i \in \tb$ and $P_i \subseteq \tb'$ \label{alg:pin:loop:begin}}{
      $\phi = lab(a_i) \wedge \bigwedge_{p \in P_i} lab(p)$ \\
      \uIf{$q_i \not \in \tb'$}{
        $\tb' := \tb' \cup \{ q_i \}$\\
        $lab(q_i) := \phi$\\
      }
      \Else{
        $\psi = lab(q_i)$ \\
        \If{$\psi \vee \phi \not \equiv \psi$}{
          $lab(q_i) := \psi \vee \phi$
        }
      }
      \label{alg:pin:loop:end}
  }
  \Return{$(\tb', lab)$}
  \caption{\textsc{AllMinAs}($\tb$)}
  \label{alg:pin}
\end{algorithm}
}
\end{frame}

\begin{frame}{Labeling Algorithm}
\begin{itemize}
\item We obtain $\tb'$ like before, but additionally a labeling $lab$.
\item All $\subseteq$-minimal valuations that satisfy $lab(A \subsume B)$ correspond to a MinA for $\tb$ wrt.$A \subsume B$ \cite[Thm.2]{orig2}. 
\item The algorithm runs in time exponential in the size of the input TBox\footnote{direct argumentation in \cite[Sec.3]{orig2}}, exhibited by $$\tb_n := \big\{ B_{i-1} \subsume P_i \sqcap Q_i, \quad P_i \subsume B_i, \quad Q_i \subsume B_i \quad  \mid \quad 1 \geq i \geq n \big\}$$

which yields $2^n$ MinAs for $\tb_n$ wrt.$B_0 \subsume B_n$ \footnote{cf. \cite[Example 1]{orig2}}.
\end{itemize}
\end{frame}

\subsection{Pinpointing via Subsumption as Black-Box}

\begin{frame}
\vspace{10mm}
\begin{algorithm}[H]
  \KwIn{A TBox $\tb = \{a_1, \ldots, a_n\}$ over $\nc$ and $A, B \in \nc$.}
  \KwOut{If $A \subsume_\tb B$ holds, one MinA for $\tb$ wrt. $A \subsume B$, else $\emptyset$.}
  \If{$A \not \subsume_\tb B$}{
    \Return{$\emptyset$}
  }
  $\mathcal{S} := \tb$ \\
  \ForEach{$a_i \in \tb$ \label{alg:linone:loop:begin}}{
  \If{$A \subsume_{\mathcal{S} \setminus \{a_i\}} B$}{
    $\mathcal{S} := \mathcal{S} \setminus \{ a_i \}$ \\
  }
  \label{alg:linone:loop:end}
  }
  \Return{$\mathcal{S}$}
  \caption{\textsc{LinOneMinA}($\tb$, $A$, $B$) (cf.\cite[Alg.1]{orig2,debugging}, \cite[Alg.7, p.97]{meng-phd})}
  \label{alg:linone}
\end{algorithm}
\end{frame}

\begin{frame}{Pinpointing via Subsumption as Black-Box}

\begin{itemize}
\item Linearily scans $\tb$ with one call to subsumption per axiom. Thus, runs in polynomial time overall. \cite[Thm.6]{orig2}
\item \enquote{\ldotsdid not terminate on \snomed in 48hrs \ldots} \cite[p.97]{meng-phd}.
\item Can be improved by using a \enquote{sliding window} approach or binary search.
\item Black-Box algorithms that compute all MinAs are certainly possible.
\end{itemize}

\end{frame}

\section{Complexity and Tradeoffs in Practice}

\begin{frame}{Complexity}
\begin{enumerate}
\item Computing all MinAs takes exponential time.\\ Is there an \alert{output polynomial algorithm}?

\item Computing one MinA takes polynomial time.\\
This is still bad for large knowledge bases.\\
Can we make trade-offs to \alert{be faster in practice}?
\end{enumerate}
\end{frame}

\begin{frame}{All MinAs}
An output polynomial algorithm?\\
\begin{itemize}
\item \cite[Thm.5]{orig2} shows this is \alert{not possible} for the case of a TBox with non-refutable part (unless P = NP).
\item In \cite[Thm.2]{hard} computing all MinAs is established to be \alert{as least as hard as computing the set of all minimal transversals of a hypergraph}, which is in coNP and no output polynomial alorithm is known (cf.\cite{hyperai}).
\end{itemize}
Also, computing properties wrt.all MinAs cannot be achieved in polynomial time (unless P = NP) \cite[Sec.4]{orig2}.
\end{frame}

\begin{frame}{One MinA}
Polynomial, but still too slow in practice.

\begin{enumerate}
\item Let's take some $\tb' \subset \tb$ with $A \subsume_{\tb'} B$ and run the algorithm on that!
\item To get $\tb'$, take the labeling algorithm and drop the re-labeling branch.
\item Other greedy algorithms might perform well/better.
\end{enumerate}
Results: 10min vs.7hrs with just 2.59\% difference in size.\footnote{simplified, cf.\cite[Sec.5]{orig2}}
\end{frame}

\begin{frame}{Recap}
\tableofcontents
\end{frame}

\begin{frame}[standout]
Questions, please!
\end{frame}

\begin{frame}[allowframebreaks]{References}

{\tiny{\bibliography{ref} \bibliographystyle{abbrv}}}
\end{frame}

\end{document}
